\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated manuscript, checked lines in response to reviewer's document

In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

Due to a variety of causes, the number of observations during a study can be different between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}}. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}. However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}}. And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .

@@

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and missing observations are explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

@@

However, it is worth reminding that ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power. Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.

@@

where \(\eta_{ijt}\) is the linear predictor for each observation \(i\) in each group \(j\), at each timepoint \(t\), and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group for the linear mean from Equation (\eqref{eq:linear-model}). Finally, \(time_{t}\) represents the covariates from each subject in each group at each time point.

@@

where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)). The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}. Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.

@@

Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness. The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}. A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}.

In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}. Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).

@@

Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present. A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline. The second derivative of the spline is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature. As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting. In other words, a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.

With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}. In Figure \ref{fig:basis-plot}A the four basis functions (and the intercept) are shown. Each of the five basis functions is evaluated at six different points (because there are six points on the timeline). The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model. The penalized coefficient estimates fitted are shown in Figure \ref{fig:basis-plot}B.

To get the weighed basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions. Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one). On the other hand, the third basis function has roughly doubled its contribution to the smooth term. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values which appear as points.

@@

To understand across the function coverage, recall that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process. Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie. In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}), if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%). In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI. As a consequence, some areas of CI for the function must have more than nominal coverage and some areas less than the nominal coverage.

@@

In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}. What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.

@@

The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}. A brief guide for model selection and diagnostics appears in Appendix A, and the code for the simulated data and figures can be found in Appendix B.

@@

This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using five basis functions. The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console). Finally, the parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, as we have indicated above.

@@

Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends. To exemplify this, consider the random deletion of 40\% of the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}A. If the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted to this data with missing observations, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D). Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.

Additionally, note that in Figure \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions). The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, they are wider than the across the function CI, resulting in the wide shaded regions. For the dataset with missing observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset. However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.

@@

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.

@@

The absence of a \emph{p-value} in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.

With this expectation of different trends in each group, computing the difference between the trends will identify if and \emph{when} the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group). Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere \emph{p-value} because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain \emph{the whole function} at a nominal value (say, 95\%).

Consider the calculation of pairwise differences for the fitted smooths that appear in Figure \ref{fig:sim-smooth-plot}B, D. Figure \ref{fig:plot-pairwise-comp} shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference. Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared. Notice that because we have included the means of each group, there is correspondence between the scale of the original data and the scale of the pairwise comparisons. This can be seen in Figure \ref{fig:sim-smooth-plot}B, where at day 5 there is essentially a difference of 50\% between \(\mbox{StO}_2\) in both groups, which corresponds to the -50\% difference in \ref{fig:plot-pairwise-comp}A. However, if there are multiple parametric terms in the model (more factors that need to be specified), such inclusion of the means can become problematic. However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.

In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other. Notice that the shaded region between days 1 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day \(\approx\) 3 it is the ``Treatment'' group that statistically on average has greater \(\mbox{StO}_2\). This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model. Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.

On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap zero for the first two days there is statistically significant no difference between the groups. However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes statistically significant as in the full dataset smooth pairwise comparison.

@@

Biomedical longitudinal data is particularly challenging to analyze due to the frequency of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA. Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}. When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error''. This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.

@@

However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs where additionally there is no biological interpretation for such effect. In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.

@@

An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF. A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.

>>>>

In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting incomplete observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when incomplete observations exist (with 40\% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

Due to a variety of causes, the number of observations during a study can be different between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with incomplete observations from the analysis\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}}. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate incomplete observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}. However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}}. And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .

@@

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and complete observations are explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

@@

However, it is worth reminding that ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of incomplete data, such as sampling bias or a drastic reduction in statistical power. Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.

@@

where \(\eta_{ijt}\) is the linear predictor for each observation \(i\) in each group \(j\), at each timepoint \(t\). Following the notation from Equation \eqref{eq:linear-model} the model parameters for each group are \(\beta_0\) (the intercept), \(\beta_{1}\), \(\beta_2\), and \(\beta_3\); \(time_{t}\) represents the covariates from each subject in each group at each time point, \(treatment_j\) represents the different treatment levels, and \(time_t \times treatment_j\) represents their interaction.

@@

where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)). The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}). Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.

@@

Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness. The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}. A solution to the knot placement limitation of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}. In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}. Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).

@@

Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present. A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline smooth. The second derivative of the spline smooth is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature. As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting. In other words, a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.

With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down trend) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}. In Figure \ref{fig:basis-plot}A the four basis functions (and the intercept) are shown. Each of the five basis functions is evaluated at six different points (because there are six points on the timeline). The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model. The penalized coefficient estimates fitted are shown in Figure \ref{fig:basis-plot}B.

To get the weighed basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one). On the other hand, the third basis function has roughly doubled its contribution to the smooth term. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values which appear as points.

@@

To understand across the function coverage, recall that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process. Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie. In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}), if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%). In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI. As a consequence, some areas of the CI for the function have more than nominal coverage and some areas less than the nominal coverage.

@@

In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}. What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their estimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.

@@

The previous sections provided the basic understanding of the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}. A brief guide for model selection and diagnostics appears in Appendix A, and the code for the simulated data and figures can be found in Appendix B.

@@

This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from Appendix A) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus intercept). The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console). Finally, the parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, as we have indicated above.

@@

Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where unbalanced data exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are incomplete. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends. To exemplify this, consider the random deletion of 40\% of the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}A. If the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted to this data with unbalanced observations, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D). Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.

Additionally, note that in Figure \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions). The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, they are wider than the across the function CI, resulting in the wide shaded regions. For the dataset with incomplete observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset. However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.

@@

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in \passthrough{\lstinline!gam\_02!}) can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.

@@

The absence of a \emph{p-value} in this context might seem odd, but the pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.

With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group). Therefore, the computation of both the difference between the smooths and the corresponding simultaneous empirical Bayesian CI around this difference is able to provide an estimation of when and by how much there is a significant difference between the different groups. Additionally, the correction provided by the simultaneous empirical Bayesian CI makes the estimation robust as we know that on average, the simultaneous CI will contain the \emph{whole function} at a nominal value (say, 95\%).

To visualize this, consider the calculation of pairwise differences for the fitted smooths that appear in Figure \ref{fig:sim-smooth-plot}B, D. Figure \ref{fig:plot-pairwise-comp} shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference. Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared. Notice that because we have included the means of each group, there is correspondence between the scale of the original data and the scale of the pairwise comparisons. This can be seen in Figure \ref{fig:sim-smooth-plot}B, where at day 5 there is essentially a difference of 50\% between \(\mbox{StO}_2\) in both groups, which corresponds to the -50\% difference in \ref{fig:plot-pairwise-comp}A. However, if there are multiple parametric terms in the model (more factors that need to be specified), such inclusion of the means can become problematic. However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.

In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where the CI does not cover 0) indicate the time interval where each group has a different mean effect than the other. Notice that the shaded region between days 1 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day \(\approx\) 3 it is the ``Treatment'' group that statistically on average has greater \(\mbox{StO}_2\). This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model. Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.

On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap zero for the first two days there is no statistically significant difference between the groups. However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison.

@@

Biomedical longitudinal data is particularly challenging to analyze due to the frequency of incomplete observations and different correlation structures in the data, which limit the use of rm-ANOVA. Although LMEMs can handle unbalanced observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}. When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error''. This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.

@@

However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data. The use of basis functions is a major advantage over models where the user has to know the non-linear relationship \emph{a priori}, such as in the case of polynomial effects in LMEMs. In addition, there is no biological interpretation for such polynomial assumptions. In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.

@@

Two Appendices which contain a basic workflow to implement GAMs in R and all the code used to create this manuscript are available as PDFs in the Supplementary Material. A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.
